---
title: "FinalProject"
author: "Ajay Karatam and Michael Rothschilds"
date: "5/13/2020"
output: html_document
---
## An Analysis of U.S Hospital Bed Capacity During the COVID-19 Pandemic
by Michael Rothschilds and Ajay Karatam

## Introduction: 
The COVID-19 pandemic has altered the lives of millions over the past few months. People have practiced Social Distancing at an unprecedented level with the intention of keeping people safe and "flattening the curve". The purpose of "flattening the curve" is to prevent hospitals from being overun with too many COVID-19 patients at once. In order to stop this from happening, public health and government officials need to project the number of cases, the rate of hospitilization, and the resulting number of hospital and ICU beds needed at any given time across all areas of the United States. The following data was collected from a team of researchers at the Harvard Global Data Institute. They made projections for different infection rates and used surveys conducted by The American Hospital Association to predict the true availability of hospital and ICU beds.

## Examine the original dataset here: 
https://www.kaggle.com/mrmorj/hospital-bed-capacity-and-covid19

## R Libraries Used
``` {r libraries}
library(tidyverse)
library(lubridate)
```



## Dataframe Setup and Tidying

```{r setup}
csv_file <- "HRR_Scorecard.csv"
hcb <- read_csv(csv_file)

df <- hcb %>%
  
  
  
  
```

